---
title: "Randomization"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Randomization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(dplyr)
library(ggplot2)
```

Here is the actual code used to perform the randomization. 

To recreate the process, we use only the columns available in the original
csv file. 

```{r}
fields <- RuckmanMeta::fields %>%
  select(field,block,whole_plot,area_acre,manure,x,y)
```

##

The design of the experiment is a split-plot design within blocks augmented
with a non-randomized study. 

## Randomization

The randomization has two steps:

1. Randomly assign harvest year to the whole plot
2. Randomly assign treatments to the fields within the whole plot

```{r}
#################################################
### Randomization
#################################################
set.seed(38) # For reproducibility

fields <- fields %>%

  # Randomize initial harvest
  group_by(block) %>%
  mutate(harvest = as.integer(ifelse(whole_plot == sample(whole_plot,1), 2019, 2020)),
         whole_plot = paste0(block, ifelse(harvest==2019, "a", "b"))) %>% # Rename whole_plot to coincide with harvest year

  # Randomize treatment
  group_by(block,whole_plot,manure) %>%
  mutate(trt = sample(c("FB","LoDiv","HiDiv")),
         trt = ifelse(manure=="no", "HiDiv", trt)) %>% # non-randomization augmentation
  ungroup()
```

We had two criteria to satisfy for a successful randomization:

- Fields 13a and 13b must have exactly one FB treatment between the two of them
- Mean and SD of field sizes in each treatment are approximately the same

```{r}
# Check to ensure fields 13a and 13b have exactly one FB treatment
fields %>%
  filter(field %in% c("13a","13b"))

# Check field sizes for each treatment-manure combination to ensure the
# mean and sd are approximately the same
fields %>%
  group_by(trt,manure) %>%
  summarize(n = n(),
            mean = mean(area_acre),
            sd   = sd(area_acre)) %>%
  arrange(manure, trt)
```

An iterative process was undertaken, updating the `set.seed` value until these
criteria were satisfied. 
The final randomization was 

```{r}
fields %>% as.data.frame
```

```{r}
ggplot(fields, aes(x=x, y=y, label=harvest)) +
  geom_text() +
  theme_bw() +
  xlim(c(0,11)) +
  ylim(c(0,14))

ggplot(fields, aes(x=x, y=y, label=trt)) +
  geom_text() +
  theme_bw() +
  xlim(c(0,11)) +
  ylim(c(0,14))
```
